Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2015-11-13 docs/ms_sageAbundance.Rmd
edits to Andy's adits

The ideal tool for forecasting future population states under climate change would be a large-scale dynamic population model [@Schurr2012a; @Merow2014]; however, developing useful models at this scale has been limited by the lack of time series data at large spatial extents and the lack of statistical machinary needed to fit high-dimensional spatial models.
Fortunately, new advances in remote sensing and statistics now allow us to overcome both of these limitations.
First, new remote-sensing (RS) methods are now producing accurate time series of species-specific plant cover at landscape scales.  These data can be fit with dynamic population models which include yearly fluctuation in climate as covariates.
Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques that overcome the dfficulties of fitting high-dimension spatiotemporal models [see @Conn2015 for a review]. These new statistical methods have yet to be applied to RS derived plant populations data at large scales.

@@

Sagebrush (*Artemisia tridentata*) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].  SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015]. Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014]. In warmer parts of its range, this could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014]. In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014]. A large-scale spatially explicit population model for sagebrush driven by interannual climate would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.

(Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].) <-- Too much detail? could move to discussion or cite as evidence that "sagebrush is sensitive to climate" above?

Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data, we demonstrate how large-scale plant population models could be used to predict how climate change will affect sagebrush and other species in the future.  Our approach relies on a process model motivated by Gompertz density-dependent population growth and a cutting-edge remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012]. We rigorously account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.

The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the largest potential.
In those cases, our approach is particularly well-suited for combining the information of population models (e.g., population status) and the spatial extent of species distribution models.

@@

The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2) which agrees with previous research showing that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013]. The posterior distribution for the effect of spring temperature on sagebrush cover broadly overlapped zero but was slightly positive (Fig. 2).  Warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].  As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.

As expected given the relatively weak climate effects in this study area, our model projects only small changes in sagebrush cover in this area with future climate change.  Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs. 4,5).
The forecasted increases are primarily driven by a positive response to precipitation (Fig. 3), but weak effects of temperature also influenced results.
A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline in the future [@Homer2015].
The discrepancy between Homer et al.'s (2015) results and ours reflects a differences in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
The Homer et al. (2015) study also found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
Thus, the decline in precipitation led to a decline in sagebrush cover.
Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of our forecasts when all sources of uncertainty are considered (Fig. 5).

(not sure this next paragraph is needed--it also sounds a bit too prescriptive about climate projection which is probably outside the realm of the paper) 
There is no clear answer to what climate model to use when making ecological forecasts. Much like weather forecasters [@Bauer2015], ecological forecasters may need to apply expert judgement to choose climate projections.
If a certain GCM tends to better predict weather in a specific study area, then that GCM may be given more weight than others.

@@

While we presented projections for sagebrush cover at the end of this century, one important feature of our aproach is that it could also produce short-term forecasts for sagebrush cover as well.  For instance this apprach could be used to model the effects of a multi-year regional drought on this species (cite Debinski et al. 2010).  Validating spatial population models against short-term predictions would give modelers a way to assess the long-term forecasts produced by the model.  It would also provide opportunities for refinement of the model itself.  This cycle of prediction, validation and refinement is missing from most population-level forecasts of the effects of climate change currently produced [@Petchey2015 prediction paper -- is there a better reference for this idea?].

Conclusion
----------
We have introduced a new approach to fitting and simulating population models at large spatial extents with plant population data derived from state of the art remote sensing.
As more species-level remote sensing datasets become available and computing power increases this approach will be applicable to a wider number of species and even larger spatial extents. Future modeling could include the effects of non-climate drivers--including the effects of species interactions and disturbance [@Clark2014].  For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford and Lauenroth  2009 ].  Fortunately, our spatio-temporal modeling framework could easily be extended to model additional species and dynamic processes as the data become availabe. The approach we have developed here fills an important gap in spatial scales between species distribution models and local-scale demographic population models.

>>>>

The ideal tool for forecasting future population states under climate change would be a large-scale dynamic population model [@Schurr2012a; @Merow2014]; however, developing useful models at this scale has been limited by the lack of time series data at large spatial extents and the lack of statistical machinary needed to fit high-dimensional spatial models.
Fortunately, new advances in remote sensing and statistics now allow us to overcome both of these limitations.
First, new remote-sensing (RS) methods are now producing accurate time series of species-specific plant cover at landscape scales.  These data can be fit with dynamic population models which include yearly fluctuation in climate as covariates.
Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques that overcome the dfficulties of fitting high-dimension spatiotemporal models [see @Conn2015 for a review]. 
These new statistical methods have yet to be applied to RS-derived plant populations data at large scales.

@@

Sagebrush (*Artemisia tridentata*) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].  
SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015]. 
Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014]. 
In warmer parts of its range, this could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014]. 
In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].
A large-scale spatially-explicit population model for sagebrush driven by interannual climate would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.

Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data, we demonstrate how large-scale plant population models could be used to predict how climate change will affect sagebrush and other species in the future. 
Our approach relies on a process model motivated by Gompertz density-dependent population growth and a cutting-edge remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012]. 
We rigorously account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the largest potential.
In those cases, our approach is particularly well-suited for combining the information of population models (e.g., population status) and the spatial extent of species distribution models.

@@

The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2), which agrees with previous research showing that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013]. 
The posterior distribution for the effect of spring temperature on sagebrush cover broadly overlapped zero but was slightly positive (Fig. 2).  
Warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].  
As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.

As expected given the relatively weak climate effects in this study area, our model projects only small changes in sagebrush cover in this area with future climate change. 
Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs. 4,5).
The forecasted increases are primarily driven by a positive response to precipitation (Fig. 3), but weak effects of temperature also influenced results.
A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline in the future [@Homer2015].
The discrepancy between Homer et al.'s (2015) results and ours reflects a difference in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
The Homer et al. (2015) study also found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
Thus, the decline in precipitation led to a decline in sagebrush cover.
Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of our forecasts when all sources of uncertainty are considered (Fig. 5).

There is no clear answer to what climate model to use when making ecological forecasts. Much like weather forecasters [@Bauer2015], ecological forecasters may need to apply expert judgement to choose climate projections.
If a certain GCM tends to better predict weather in a specific study area, then that GCM may be given more weight than others.
Extensions of these ideas are the subject of ongoing research.

@@

While we presented projections for sagebrush cover at the end of this century, one important feature of our aproach is that it could also produce short-term forecasts for sagebrush cover as well.  
For instance this apprach could be used to model the effects of a multi-year regional drought on this species (cite Debinski et al. 2010).  
Validating spatial population models against short-term predictions would give modelers a way to assess the long-term forecasts produced by the model.  
It would also provide opportunities for refinement of the model itself.  
This cycle of prediction, validation and refinement is missing from most population-level forecasts of the effects of climate change currently produced [@Petchey2015].

Conclusion
----------
We have introduced a new approach to fitting and simulating population models at large spatial extents with plant population data derived from state of the art remote sensing.
As more species-level remote sensing datasets become available and computing power increases this approach will be applicable to a wider number of species and even larger spatial extents.
Future modeling could include the effects of non-climate drivers--including the effects of species interactions and disturbance [@Clark2014].  
For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford2009].  
Fortunately, our spatio-temporal modeling framework could easily be extended to model additional species and dynamic processes as the data become availabe. 
The approach we have developed here fills an important gap in spatial scales between species distribution models and local-scale demographic population models.
